---
title: "5_self_admin_data"
author: "nicholas ruiz"
date: "2026-08-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(writexl)
library(lme4)
library(lmerTest)
library(sjPlot)

inc_sneek <- read.csv("data/inc_sneek.csv")
```


# ------------------------------------------------------------
# 1. Final-analysis animal IDs
# ------------------------------------------------------------

```{r}
animal_ids <- read_excel("data/animal_ids_corrected.xlsx",
                         col_types = "text"
) %>%
  transmute(
    subject = str_squish(subject),
    reward  = str_to_lower(str_squish(reward)),
    sex     = str_to_lower(str_squish(sex)),
    day     = if_else(parse_number(day) == 1, 1L, 30L)
  )
```

# ------------------------------------------------------------
# 2. Find all Combined Incubation workbooks
# ------------------------------------------------------------

```{r}
incubation_files <- list.files(
  path = "data",
  pattern = "^Combined incubation data_.*\\.xlsx$",
  full.names = TRUE,
  ignore.case = TRUE
)
```

# ------------------------------------------------------------
# 3. Function for IVSA acquisition sheets
#    Infusions, active lever, inactive lever
# ------------------------------------------------------------

```{r}
read_ivsa <- function(file, sheet) {
  
  x <- read_excel(
    file,
    sheet = sheet,
    col_names = FALSE,
    col_types = "text"
  )
  
  # Find the row containing "ANIMAL ID"
  header_row <- which(
    apply(x, 1, \(z)
          any(
            str_to_upper(str_squish(z)) == "ANIMAL ID",
            na.rm = TRUE
          )
    )
  )[1]
  
  header <- x[header_row, ] %>%
    unlist(use.names = FALSE) %>%
    str_squish() %>%
    str_to_upper()
  
  # Locate animal ID and the 10 acquisition days/nights
  id_col <- which(header == "ANIMAL ID")[1]
  
  day_cols <- which(
    str_detect(
      header,
      "^(DAY|NIGHT)\\s*(?:[1-9]|10)$"
    )
  )
  
  day_nums <- parse_number(header[day_cols])
  
  # Retain only ID + the acquisition data
  x <- x[
    (header_row + 1):nrow(x),
    c(id_col, day_cols)
  ]
  
  names(x) <- c(
    "subject",
    paste0("day_", day_nums)
  )
  
  reward <- str_extract(
    sheet,
    regex("heroin|sucrose", ignore_case = TRUE)
  ) %>%
    str_to_lower()
  
  sex <- str_extract(
    sheet,
    regex("female|male", ignore_case = TRUE)
  ) %>%
    str_to_lower()
  
  # Important: test "inactive" before "active"
  behavior <- case_when(
    str_detect(sheet, regex("inactive", ignore_case = TRUE)) ~ "inactive",
    str_detect(sheet, regex("active",   ignore_case = TRUE)) ~ "active",
    TRUE                                                    ~ "infusion"
  )
  
  x %>%
    mutate(
      subject = str_squish(subject)
    ) %>%
    pivot_longer(
      -subject,
      names_to = "day",
      names_prefix = "day_",
      values_to = "value"
    ) %>%
    mutate(
      reward = reward,
      sex = sex,
      phase = "ivsa",
      day = as.integer(day),
      behavior = behavior
    ) %>%
    
    # Filters to final-analysis animals without duplicating
    # animals that have both cue-reactivity days
    semi_join(
      animal_ids,
      by = c("subject", "reward", "sex")
    ) %>%
    
    mutate(
      value = parse_double(value)
    ) %>%
    
    select(
      subject, reward, sex,
      phase, day, behavior, value
    )
}
```

# ------------------------------------------------------------
# 4. Function for cue-reactivity sheets
# ------------------------------------------------------------

```{r}
read_cue <- function(file, sheet) {
  
  # The first six columns contain the cohort-by-cohort raw data.
  # The additional tables to the right in some heroin sheets
  # duplicate these data, so we deliberately ignore them.
  x <- read_excel(
    file,
    sheet = sheet,
    col_names = FALSE,
    col_types = "text"
  ) %>%
    select(1:6)
  
  names(x) <- c(
    "subject",
    "box",
    "condition",
    "infusion",
    "active",
    "inactive"
  )
  
  reward <- str_extract(
    sheet,
    regex("heroin|sucrose", ignore_case = TRUE)
  ) %>%
    str_to_lower()
  
  sex <- str_extract(
    sheet,
    regex("female|male", ignore_case = TRUE)
  ) %>%
    str_to_lower()
  
  x %>%
    mutate(
      
      # Identify Day 1 vs later incubation test
      day = case_when(
        str_detect(
          subject,
          regex("^day\\s*1\\b", ignore_case = TRUE)
        ) ~ 1L,
        
        # Day 30, 31, 32, 33, 34 etc. are treated as Day 30
        str_detect(
          subject,
          regex("^day\\s*3\\d\\b", ignore_case = TRUE)
        ) ~ 30L,
        
        TRUE ~ NA_integer_
      )
    ) %>%
    
    # Carry the cue-test day down to the animal rows
    fill(day) %>%
    
    mutate(
      subject = str_squish(subject),
      reward = reward,
      sex = sex,
      phase = "cue_reactivity"
    ) %>%
    
    # This join includes ONLY the requested cue-test day(s)
    # for each final-analysis animal
    semi_join(
      animal_ids,
      by = c("subject", "reward", "sex", "day")
    ) %>%
    
    pivot_longer(
      c(infusion, active, inactive),
      names_to = "behavior",
      values_to = "value"
    ) %>%
    
    mutate(
      value = parse_double(value)
    ) %>%
    
    select(
      subject, reward, sex,
      phase, day, behavior, value
    )
}
```

# ------------------------------------------------------------
# 5. Import everything into ONE tidy dataframe
# ------------------------------------------------------------

```{r}
behavior_raw <- incubation_files %>%
  map(\(file) {
    
    excel_sheets(file) %>%
      
      map(\(sheet) {
        
        if (
          str_detect(
            sheet,
            regex("cue reactivity", ignore_case = TRUE)
          )
        ) {
          
          read_cue(file, sheet)
          
        } else if (
          str_detect(
            sheet,
            regex(
              "infusion|active lever|inactive lever",
              ignore_case = TRUE
            )
          )
        ) {
          
          read_ivsa(file, sheet)
          
        } else {
          
          NULL
        }
      }) %>%
      
      compact() %>%
      list_rbind()
  }) %>%
  
  list_rbind() %>%
  
  arrange(
    reward, sex, subject,
    phase, day, behavior
  )
```

# Run this section after the existing code has created `animal_ids` and
# `behavior_raw`. All files are assumed to be in data_dir.

```{r}
data_dir <- "data/"

missing_ids <- read_excel(
  file.path(data_dir, "missing_animal_ids.xlsx"),
  sheet = "missing_by_row",
  col_types = "text"
) %>%
  transmute(
    subject = str_extract(str_to_upper(str_squish(subject)), "[A-Z]*\\d+"),
    reward = str_to_lower(str_squish(reward)),
    sex = str_to_lower(str_squish(sex)),
    day = if_else(parse_number(day) == 1, 1L, 30L),
    across(
      c(
        missing_infusion,
        missing_active,
        missing_inactive,
        missing_cue
      ),
      ~ str_to_lower(.x) == "true"
    )
  )

trap_files <- file.path(
  data_dir,
  c(
    "Heroin Incubation + D1 TRAP data 2021.xlsx",
    "Heroin Incubation + D2 TRAP data 2021.xlsx"
  )
)

clean_subject <- function(x) {
  str_extract(
    str_to_upper(str_squish(as.character(x))),
    "[A-Z]*\\d+"
  )
}
```

# ---------------------------------------------------------------------------
# Read the vertically stacked infusion/active/inactive tables in one cohort tab
# ---------------------------------------------------------------------------

```{r}
read_trap_ivsa <- function(file, sheet) {

  x <- read_excel(
    file,
    sheet = sheet,
    col_names = FALSE,
    col_types = "text"
  )

  labels <- x[[1]] %>%
    str_squish() %>%
    str_to_upper()

  table_rows <- which(
    labels == "INFUSIONS" |
      str_detect(labels, "^ACTIVE\\s+LEVER") |
      str_detect(labels, "^INACTIVE\\s+LEVER")
  )

  map(table_rows, \(header_row) {

    behavior <- case_when(
      str_detect(labels[header_row], "^INACTIVE\\s+LEVER") ~ "inactive",
      str_detect(labels[header_row], "^ACTIVE\\s+LEVER")   ~ "active",
      TRUE                                                  ~ "infusion"
    )

    first_data_row <- header_row + 2L

    # Animal rows form one contiguous block below each table header.
    block_ids <- clean_subject(x[[1]][first_data_row:nrow(x)])
    first_blank <- which(is.na(block_ids))[1]

    last_data_row <- if (is.na(first_blank)) {
      nrow(x)
    } else {
      first_data_row + first_blank - 2L
    }

    data_rows <- first_data_row:last_data_row

    header <- x[header_row, ] %>%
      unlist(use.names = FALSE) %>%
      str_squish() %>%
      str_to_upper()

    candidate_cols <- which(
      str_detect(header, "^DAY\\s*(?:[1-9]|10|OFF)$")
    )

    # Most tabs have 11 dated columns because one is a day off. In one cohort,
    # the DAY OFF label is shifted by one column. Select the 10 populated
    # acquisition columns and number them chronologically rather than trusting
    # that label.
    populated <- map_int(
      candidate_cols,
      \(column) sum(
        !is.na(suppressWarnings(parse_double(x[[column]][data_rows])))
      )
    )

    day_cols <- candidate_cols[
      order(populated, decreasing = TRUE)[seq_len(min(10L, length(candidate_cols)))]
    ] %>%
      sort()

    if (length(day_cols) != 10L) {
      stop("Could not identify 10 IVSA days in sheet: ", sheet)
    }

    out <- x[data_rows, c(1, 3, day_cols)]
    names(out) <- c(
      "subject",
      "condition",
      paste0("day_", 1:10)
    )

    needed <- missing_ids %>%
      filter(.data[[paste0("missing_", behavior)]]) %>%
      distinct(subject, reward, sex)

    out %>%
      mutate(
        subject = clean_subject(subject),
        condition = str_to_lower(str_squish(condition)),
        reward = "heroin"
      ) %>%
      filter(condition == "heroin") %>%
      inner_join(needed, by = c("subject", "reward")) %>%
      pivot_longer(
        starts_with("day_"),
        names_to = "day",
        names_prefix = "day_",
        values_to = "value"
      ) %>%
      transmute(
        subject,
        reward,
        sex,
        phase = "ivsa",
        day = as.integer(day),
        behavior,
        value = parse_double(value)
      )
  }) %>%
    list_rbind()
}
```

# ---------------------------------------------------------------------------
# Read repeated Day 1/Day 30 blocks in a consolidated cue-reactivity tab
# ---------------------------------------------------------------------------

```{r}
read_trap_cue <- function(file, sheet = "Cue reactivity") {

  x <- read_excel(
    file,
    sheet = sheet,
    col_names = FALSE,
    col_types = "text"
  )

  row_text <- apply(
    x,
    1,
    \(z) str_c(z[!is.na(z)], collapse = " | ")
  )

  header_rows <- which(
    map_lgl(
      seq_len(nrow(x)),
      \(row) any(
        str_to_upper(str_squish(unlist(x[row, ], use.names = FALSE))) ==
          "ANIMAL ID",
        na.rm = TRUE
      )
    )
  )

  # Determine the day attached to each repeated header. One D1 workbook block
  # has a blank Day 30 label, so a second header following Day 1 is Day 30.
  block_days <- rep(NA_integer_, length(header_rows))

  for (i in seq_along(header_rows)) {
    first_context_row <- if (i == 1L) 1L else header_rows[i - 1L] + 1L

    day_labels <- str_extract_all(
      row_text[first_context_row:header_rows[i]],
      regex("\\bDAY\\s*(?:1|3\\d)\\b", ignore_case = TRUE)
    ) %>%
      unlist()

    block_days[i] <- if (length(day_labels) > 0L) {
      if_else(parse_number(last(day_labels)) == 1, 1L, 30L)
    } else if (i > 1L && block_days[i - 1L] == 1L) {
      30L
    } else {
      NA_integer_
    }
  }

  map2(header_rows, block_days, \(header_row, cue_day) {

    header <- x[header_row, ] %>%
      unlist(use.names = FALSE) %>%
      str_squish() %>%
      str_to_upper()

    id_col <- which(header == "ANIMAL ID")[1]
    condition_col <- which(header == "CONDITION")[1]
    infusion_col <- which(header == "INFUSIONS")[1]
    active_col <- which(header == "ACTIVE LEVER")[1]
    inactive_col <- which(header == "INACTIVE LEVER")[1]

    first_data_row <- header_row + 1L
    block_ids <- clean_subject(x[[id_col]][first_data_row:nrow(x)])
    first_blank <- which(is.na(block_ids))[1]

    last_data_row <- if (is.na(first_blank)) {
      nrow(x)
    } else {
      first_data_row + first_blank - 2L
    }

    out <- x[
      first_data_row:last_data_row,
      c(id_col, condition_col, infusion_col, active_col, inactive_col)
    ]

    names(out) <- c(
      "subject",
      "condition",
      "infusion",
      "active",
      "inactive"
    )

    needed <- missing_ids %>%
      filter(missing_cue, day == cue_day) %>%
      distinct(subject, reward, sex, day)

    out %>%
      mutate(
        subject = clean_subject(subject),
        condition = str_to_lower(str_squish(condition)),
        reward = "heroin",
        day = cue_day
      ) %>%
      filter(condition == "heroin") %>%
      inner_join(
        needed,
        by = c("subject", "reward", "day")
      ) %>%
      pivot_longer(
        c(infusion, active, inactive),
        names_to = "behavior",
        values_to = "value"
      ) %>%
      transmute(
        subject,
        reward,
        sex,
        phase = "cue_reactivity",
        day,
        behavior,
        value = parse_double(value)
      )
  }) %>%
    list_rbind()
}
```

# ---------------------------------------------------------------------------
# Extract only previously missing records and append non-overlapping keys
# ---------------------------------------------------------------------------

```{r}
trap_raw <- trap_files %>%
  map(\(file) {
    cohort_sheets <- excel_sheets(file) %>%
      keep(
        ~ str_detect(
          .x,
          regex(
            "^(D1 TRAP|D2 TRAP|MIXED D1 D2 TRAP)",
            ignore_case = TRUE
          )
        )
      )

    bind_rows(
      map(cohort_sheets, \(sheet) read_trap_ivsa(file, sheet)) %>%
        list_rbind(),
      read_trap_cue(file)
    )
  }) %>%
  list_rbind()

behavior_key <- c(
  "subject", "reward", "sex", "phase", "day", "behavior"
)

if (
  trap_raw %>%
    count(across(all_of(behavior_key))) %>%
    filter(n > 1L) %>%
    nrow() > 0L
) {
  stop("The TRAP import produced duplicate subject/day/behavior keys.")
}

behavior_raw <- bind_rows(
  behavior_raw,
  trap_raw %>%
    anti_join(
      behavior_raw %>% distinct(across(all_of(behavior_key))),
      by = behavior_key
    )
) %>%
  arrange(reward, sex, subject, phase, day, behavior)
```

# ---------------------------------------------------------------------------
# Recalculate missingness from the updated behavior_raw and export the audit
# ---------------------------------------------------------------------------

```{r}
missing_audit <- animal_ids %>%
  transmute(
    subject = clean_subject(subject),
    reward = str_to_lower(str_squish(reward)),
    sex = str_to_lower(str_squish(sex)),
    day = if_else(parse_number(as.character(day)) == 1, 1L, 30L)
  ) %>%
  left_join(
    behavior_raw %>%
      filter(
        phase == "ivsa",
        behavior %in% c("infusion", "active", "inactive")
      ) %>%
      group_by(subject, reward, sex, behavior) %>%
      summarise(
        n_days = n_distinct(day[!is.na(value)]),
        .groups = "drop"
      ) %>%
      pivot_wider(
        names_from = behavior,
        values_from = n_days,
        names_prefix = "n_",
        values_fill = 0L
      ),
    by = c("subject", "reward", "sex")
  ) %>%
  left_join(
    behavior_raw %>%
      filter(
        phase == "cue_reactivity",
        behavior %in% c("infusion", "active", "inactive")
      ) %>%
      group_by(subject, reward, sex, day) %>%
      summarise(
        n_cue_behaviors = n_distinct(behavior[!is.na(value)]),
        .groups = "drop"
      ),
    by = c("subject", "reward", "sex", "day")
  ) %>%
  mutate(
    across(
      c(n_infusion, n_active, n_inactive, n_cue_behaviors),
      ~ replace_na(.x, 0L)
    ),
    missing_infusion = n_infusion < 10L,
    missing_active = n_active < 10L,
    missing_inactive = n_inactive < 10L,
    missing_cue = n_cue_behaviors < 3L,
    missing_data_types = pmap_chr(
      list(
        missing_infusion,
        missing_active,
        missing_inactive,
        missing_cue
      ),
      \(inf, act, inact, cue) {
        str_c(
          c(
            if (inf) "ivsa infusion",
            if (act) "ivsa active",
            if (inact) "ivsa inactive",
            if (cue) "cue reactivity"
          ),
          collapse = "; "
        )
      }
    )
  )

final_missing_by_row <- missing_audit %>%
  filter(
    missing_infusion |
      missing_active |
      missing_inactive |
      missing_cue
  ) %>%
  mutate(day = str_c("day ", day)) %>%
  select(
    subject, reward, sex, day,
    missing_infusion, missing_active,
    missing_inactive, missing_cue,
    missing_data_types
  )

final_missing_unique_animals <- missing_audit %>%
  group_by(subject, reward, sex) %>%
  summarise(
    days_in_animal_ids = str_c(
      "day ",
      sort(unique(day)),
      collapse = ", "
    ),
    missing_infusion = any(missing_infusion),
    missing_active = any(missing_active),
    missing_inactive = any(missing_inactive),
    missing_cue_day1 = any(day == 1L & missing_cue),
    missing_cue_day30 = any(day == 30L & missing_cue),
    missing_any = any(
      missing_infusion |
        missing_active |
        missing_inactive |
        missing_cue
    ),
    .groups = "drop"
  ) %>%
  filter(missing_any)

write_xlsx(
  list(
    final_missing_by_row = final_missing_by_row,
    final_missing_unique = final_missing_unique_animals
  ),
  file.path(data_dir, "final_missing_animal_ids.xlsx")
)
```


# ------------------------------------------------------------
# 6. plot drug by ivsa behavior
# ------------------------------------------------------------

```{r}
behavior_raw %>%
  filter(
    phase == "ivsa",
    behavior %in% c("active", "inactive", "infusion")
  ) %>%
  mutate(
    reward = factor(reward, levels = c("sucrose", "heroin")),
    behavior = factor(
      behavior,
      levels = c("active", "inactive", "infusion"),
      labels = c("Active Lever", "Inactive Lever", "Infusions")
    ),
    day = factor(day, levels = 1:10)
  ) %>%
  ggplot(
    aes(
      x = day,
      y = value,
      fill = reward
    )
  ) +
  
  # Group means
  stat_summary(
    fun = mean,
    geom = "col",
    position = position_dodge(width = 0.8),
    width = 0.7,
    alpha = 0.4
  ) +
  
  # Individual animals
  geom_point(
    aes(color = reward),
    position = position_jitterdodge(
      jitter.width = 0.15,
      dodge.width = 0.8
    ),
    size = 1.7,
    alpha = 0.15
  ) +
  
  facet_wrap(
    ~behavior,
    ncol = 1,
    scales = "free_y"
  ) +
  
  labs(
    x = "IVSA Day",
    y = "Response",
    fill = "Reward",
    color = "Reward"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  ) -> ivsa_plot

ggsave("figures/ivsa_data.png", plot = ivsa_plot)
print(ivsa_plot)
```

# ------------------------------------------------------------
# 7. summarizing ivsa data and merging w inc_sneek
# ------------------------------------------------------------

```{r}
inc_ivsa <- inc_sneek %>%
  left_join(
    
    behavior_raw %>%
      filter(
        phase == "ivsa",
        behavior %in% c("infusion", "active", "inactive")
      ) %>%
      
      mutate(
        period = if_else(day <= 5, "early", "late"),
        
        behavior = recode(
          behavior,
          infusion = "inf",
          active   = "act",
          inactive = "inact"
        )
      ) %>%
      
      group_by(subject, behavior, period) %>%
      
      summarise(
        avg = mean(value, na.rm = TRUE),
        sum = sum(value, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      
      pivot_wider(
        names_from = c(behavior, period),
        values_from = c(avg, sum),
        names_glue = "{behavior}_{.value}_{period}"
      ),
    
    by = "subject"
  )

inc_ivsa <- inc_ivsa %>% 
  mutate(inf_sum = inf_sum_early + inf_sum_late,
         act_sum = act_sum_early + act_sum_late,
         inact_sum = inact_sum_early + inact_sum_late)


# collapse across the 6 timepoints
inc_model <- inc_ivsa %>%
  group_by(subject, drug, sex, session) %>%
  summarise(
    lever = sum(lever, na.rm = TRUE),
    beam = sum(beam, na.rm = TRUE),
    sneek = sum(sneek, na.rm = TRUE),
    groom = sum(groom, na.rm = TRUE),
    shake = sum(shake, na.rm = TRUE),
    
    inf_avg_early   = first(inf_avg_early),
    inf_avg_late    = first(inf_avg_late),
    act_avg_early   = first(act_avg_early),
    act_avg_late    = first(act_avg_late),
    inact_avg_early = first(inact_avg_early),
    inact_avg_late  = first(inact_avg_late),
    
    inf_sum_early   = first(inf_sum_early),
    inf_sum_late    = first(inf_sum_late),
    act_sum_early   = first(act_sum_early),
    act_sum_late    = first(act_sum_late),
    inact_sum_early = first(inact_sum_early),
    inact_sum_late  = first(inact_sum_late),
    
    inf_sum = inf_sum_early + inf_sum_late,
    act_sum = act_sum_early + act_sum_late,
    inact_sum = inact_sum_early + inact_sum_late,
    
    .groups = "drop"
  )
```



```{r}
behavior_raw %>%
  filter(behavior == "active" & phase == "ivsa") -> active_levers

mod_act <- lmer(value ~ day * reward + (1|subject), data = active_levers)
summary(mod_act)
plot_model(mod_act, type = "pred")
plot_model(mod_act, type = "int")

behavior_raw %>%
  filter(behavior == "inactive" & phase == "ivsa") -> inactive_levers

mod_inact <- lmer(value ~ day * reward + (1|subject), data = inactive_levers)
summary(mod_inact)
plot_model(mod_inact, type = "pred")
plot_model(mod_inact, type = "int")

behavior_raw %>%
  filter(behavior == "infusion" & phase == "ivsa") -> infusions

mod_inf <- lmer(value ~ day * reward + (1|subject), data = infusions)
summary(mod_inf)
plot_model(mod_inf, type = "pred")
plot_model(mod_inf, type = "int")
```

```{r}
mod_ivsa <- lmer(lever ~ drug*session + (1|subject),data = inc_ivsa)
summary(mod_ivsa)
anova(mod_ivsa)

mod_ivsa_beh <- lmer(lever ~ session*drug + inf_sum + (1|subject),data = inc_ivsa)
summary(mod_ivsa_beh)
```


# ------------------------------------------------------------
# QC for IVSA dataset
# ------------------------------------------------------------

```{r}
ivsa <- behavior_raw %>%
  filter(
    phase == "ivsa",
    behavior %in% c("active", "inactive", "infusion"),
    !is.na(value)
  )

# Find duplicate animal/day/behavior records
duplicate_keys <- ivsa %>%
  count(subject, reward, sex, day, behavior) %>%
  filter(n > 1)

duplicate_keys

# Animals for whom more than one acquisition trajectory has been imported
ambiguous_animals <- duplicate_keys %>%
  distinct(subject, reward, sex)

ambiguous_animals

# Inspect extreme inactive-lever values
ivsa %>%
  filter(behavior == "inactive") %>%
  arrange(desc(value)) %>%
  select(subject, reward, sex, day, value) %>%
  slice_head(n = 20)

# completely remove dupes
ivsa_clean <- ivsa %>%
  anti_join(
    ambiguous_animals,
    by = c("subject", "reward", "sex")
  ) %>%
  mutate(
    subject_uid = interaction(subject, reward, sex, drop = TRUE),
    day_c = day - 1
  )

# remove a single outlier
ivsa_clean$value[602] <- NA

# rodent count by reward
ivsa_clean %>% group_by(reward) %>% summarize(n = n_distinct(subject))
```





```{r}
ivsa_lever_plot <- ivsa_clean %>%
  filter(behavior %in% c("active", "inactive")) %>%
  mutate(
    behavior = factor(
      behavior,
      levels = c("active", "inactive"),
      labels = c("Active lever", "Inactive lever")
    ),
    reward = factor(
      reward,
      levels = c("heroin", "sucrose"),
      labels = c("Heroin", "Sucrose")
    )
  ) %>%
  ggplot(aes(x = day, y = value)) +

  # Individual animals
  geom_line(
    aes(
      group = interaction(subject_uid, behavior),
      linetype = behavior
    ),
    alpha = 0.12
  ) +

  # Group means
  stat_summary(
    aes(
      group = behavior,
      linetype = behavior
    ),
    fun = mean,
    geom = "line",
    linewidth = 1.2
  ) +

  # Mean +/- SE
  stat_summary(
    aes(
      group = behavior,
      linetype = behavior
    ),
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15
  ) +

  facet_wrap(~reward, scales = "free_y") +

  scale_x_continuous(breaks = 1:10) +

  labs(
    x = "Self-administration day",
    y = "Lever presses",
    linetype = NULL
  ) +

  theme_classic(base_size = 14)

ivsa_lever_plot

ggsave(ivsa_lever_plot, file = "figures/ivsa_lever_plot.png")

ivsa_reinforcer_plot <- ivsa_clean %>%
  filter(behavior == "infusion") %>%
  mutate(
    reward = factor(
      reward,
      levels = c("heroin", "sucrose"),
      labels = c("Heroin", "Sucrose")
    )
  ) %>%
  ggplot(aes(x = day, y = value)) +

  geom_line(
    aes(group = subject_uid),
    alpha = 0.12
  ) +

  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "line",
    linewidth = 1.2
  ) +

  stat_summary(
    aes(group = 1),
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.15
  ) +

  facet_wrap(~reward, scales = "free_y") +

  scale_x_continuous(breaks = 1:10) +

  labs(
    x = "Self-administration day",
    y = "Reinforcers earned"
  ) +

  theme_classic(base_size = 14)

ivsa_reinforcer_plot

ggsave(ivsa_reinforcer_plot, file = "figures/ivsa_reinforcer_plot.png")
```

```{r}
ivsa_wide <- ivsa_clean %>%
  select(
    subject_uid, subject, reward, sex,
    day, behavior, value
  ) %>%
  pivot_wider(
    names_from = behavior,
    values_from = value
  )

acq_summary <- ivsa_wide %>%
  group_by(subject_uid, subject, reward, sex) %>%
  summarise(
    active_total = sum(active, na.rm = TRUE),
    inactive_total = sum(inactive, na.rm = TRUE),
    reinforcer_total = sum(infusion, na.rm = TRUE),

    active_late = mean(active[day >= 6], na.rm = TRUE),
    reinforcer_late = mean(infusion[day >= 6], na.rm = TRUE),

    .groups = "drop"
  )

cue_total <- inc_sneek %>%
  mutate(
    cue_day = case_when(
      session == 1 ~ 1L,
      session == 2 ~ 30L,
      TRUE ~ NA_integer_
    )
  ) %>%
  group_by(subject, cue_day) %>%
  summarise(
    cue_lever = sum(lever, na.rm = TRUE),
    .groups = "drop"
  ) %>%

  # Use animal_ids to recover reward identity
  left_join(
    animal_ids %>%
      distinct(subject, reward, day),
    by = c(
      "subject",
      "cue_day" = "day"
    )
  )

acq_cue <- cue_total %>%
  inner_join(
    acq_summary,
    by = c("subject", "reward")
  ) %>%
  mutate(
    session = factor(
      cue_day,
      levels = c(1, 30),
      labels = c("Day 1", "Day 30")
    )
  ) %>%

  # Because absolute exposure is not comparable across the two
  # protocols, standardize acquisition WITHIN reward group.
  group_by(reward) %>%
  mutate(
    z_reinforcer_total = as.numeric(scale(reinforcer_total)),
    z_active_total = as.numeric(scale(active_total)),
    z_inactive_total = as.numeric(scale(inactive_total))
  ) %>%
  ungroup()

m_sa_cue_heroin <- lmer(
  cue_lever ~ session * z_reinforcer_total +
    (1 | subject),
  data = acq_cue %>%
    filter(reward == "heroin")
)

summary(m_sa_cue_heroin)
anova(m_sa_cue_heroin)

m_sa_cue_sucrose <- lmer(
  cue_lever ~ session * z_reinforcer_total +
    (1 | subject),
  data = acq_cue %>%
    filter(reward == "sucrose")
)

summary(m_sa_cue_sucrose)
anova(m_sa_cue_sucrose)
```

```{r}
acq_cue %>%
  mutate(
    reward = factor(
      reward,
      levels = c("heroin", "sucrose"),
      labels = c("Heroin", "Sucrose")
    )
  ) %>%
  ggplot(
    aes(
      x = reinforcer_total,
      y = cue_lever,
      linetype = session
    )
  ) +

  geom_point(alpha = 0.6) +

  geom_smooth(
    aes(group = session),
    method = "lm",
    se = TRUE
  ) +

  facet_wrap(
    ~reward,
    scales = "free_x"
  ) +

  labs(
    x = "Total reinforcers earned during self-administration",
    y = "Cue-test active lever presses",
    linetype = "Cue test"
  ) +

  theme_classic(base_size = 14) -> ivsa_vs_cue_tests

ivsa_vs_cue_tests

ggsave(ivsa_vs_cue_tests, file = "figures/ivsa_vs_cue_tests.png")
```



